---
title: "Mini-Project 03 - Who Goes There? US Internal Migration and Implications for Congressional Reapportionment"
author: "khushwant-khatri590"
editor:
    mode: source
format:
    html:
        code-fold: true
---

## Introduction

In this mini-project, I explore patterns of internal migration within the 
United States, with a focus on **California**. Using data from the American 
Community Survey (ACS) — the premier national survey conducted by the US 
Census Bureau — I analyze how people move between states and metropolitan 
areas, and what those movements mean for congressional representation.

The ACS collects data continuously, providing an up-to-date picture of the 
American public. For this project, I use two types of ACS data: state-level 
ACS-1 migration flows (based on one-year samples) to track movement between 
all 50 states, and metro-level ACS-5 flows (based on five-year samples) to 
understand which cities are most connected to California's largest metro area, 
Los Angeles.

This analysis practices key data skills including downloading and parsing 
Excel and JSON files directly from government sources, combining multiple 
data sets using joins, computing summary statistics across groups, and 
building publication-quality visualizations. The ultimate goal is to forecast 
California's population in 2030, estimate its congressional apportionment 
after the next census, and propose a data-driven advertising strategy to 
help California grow its political representation in Congress.



## Data Acquisition and Preparation

In this section, I download all data needed for the analysis. I use a 
caching strategy throughout — files are only downloaded once and saved 
locally in the `data/mp03` folder to avoid redundant web requests.

### Setup: Load Libraries and Create Data Folder

```{r setup}
#| cache: false
#| message: false
#| warning: false

# Load all libraries
library(tidyverse)
library(tidycensus)
library(readxl)
library(jsonlite)
library(gt)
library(fs)
library(scales)
library(patchwork)
library(vctrs)

# Create data directory if it does not exist
data_mp03 <- fs::path("data", "mp03")
if (!fs::dir_exists(data_mp03)) {
  fs::dir_create(data_mp03, recurse = TRUE)
}

message("Setup complete!")
```

### Task 1: State Population Data

```{r task1-population}
#| cache: false
#| message: false
#| warning: false

# Helper function: download one year of state population
get_state_pop_year <- function(yr) {
  get_acs(
    variables   = c(total_population = "B01003_001"),
    geography   = "state",
    year        = yr,
    survey      = "acs1",
    geometry    = TRUE,
    cache_table = TRUE
  ) |>
    mutate(year = yr)
}

# ACS-1 was not released in 2020 due to COVID-19
available_years <- c(2015, 2016, 2017, 2018, 2019, 
                     2021, 2022, 2023, 2024)

# Download all years
state_pop_raw <- map(available_years, get_state_pop_year) |>
  list_rbind()

# Clean up
state_pop <- state_pop_raw |>
  select(
    geoid      = GEOID,
    state_name = NAME,
    population = estimate,
    year
  ) |>
  filter(!state_name %in% c(
    "Puerto Rico",
    "District of Columbia"
  ))

glimpse(state_pop)
```

The state population data contains `r nrow(state_pop)` rows covering 
50 states across `r length(available_years)` years (2015–2024, 
excluding 2020).



### Task 2: State-to-State Migration Flows — 2024

```{r task2-flows-2024}
#| cache: false 
#| message: false
#| warning: false

state_to_state_migration_2024 <- function() {

  url_2024 <- paste0(
    "https://www2.census.gov/programs-surveys/demo/tables/",
    "geographic-mobility/2024/state-to-state-migration/",
    "State_to_State_Migration_Table_2024_T13.xlsx"
  )

  dest_file <- fs::path(data_mp03, "state_to_state_2024.xlsx")

  if (!fs::file_exists(dest_file)) {
    message("Downloading 2024 migration file...")
    download.file(url_2024, destfile = dest_file, mode = "wb")
    message("Download complete.")
  } else {
    message("Using cached 2024 file.")
  }

  # 2024 file is already in long format
  # Col1 = current state, Col2 = prior state, Col3 = estimate
  raw <- read_excel(
    dest_file,
    sheet     = "Table",
    skip      = 0,
    col_names = FALSE
  )

  # Rename columns
  colnames(raw) <- c("state_current", "state_1y", "population", "moe")

  state_names_50 <- state.name

  # Cross-state flows (exclude "X" which means same state, exclude DC)
  flows_table <- raw |>
    filter(
      state_current %in% state_names_50,
      state_1y      %in% state_names_50,
      state_current != state_1y,
      population    != "X"
    ) |>
    mutate(
      population = replace(population, population == "N", "0"),
      population = suppressWarnings(as.integer(population))
    ) |>
    filter(!is.na(population)) |>
    select(state_current, state_1y, population)

  # Stayed in state rows (where state_current == state_1y, marked "X")
  # Get these from Supplemental sheet instead
  raw_supp <- read_excel(
    dest_file,
    sheet     = "Supplemental - Current Res",
    skip      = 0,
    col_names = FALSE
  )

  # Check structure
  colnames(raw_supp) <- paste0("col", 1:ncol(raw_supp))

  flows_stayed <- raw_supp |>
    filter(col1 %in% state_names_50) |>
    mutate(
      state_current = col1,
      state_1y      = col1,
      stayed = suppressWarnings(as.integer(
        str_remove_all(as.character(col2), ",")
      ))
    ) |>
    filter(!is.na(stayed)) |>
    mutate(population = stayed) |>
    select(state_current, state_1y, population)

  # Combine
  bind_rows(flows_table, flows_stayed) |>
    mutate(
      year_current = 2024L,
      year_1y      = 2023L
    ) |>
    select(state_current, state_1y, population, year_current, year_1y)
}

flows_2024 <- state_to_state_migration_2024()
message("2024 rows: ", nrow(flows_2024))
glimpse(flows_2024)
```


### Task 3: State-to-State Migration Flows — 2023

```{r task3-flows-2023}
#| cache: false 
#| message: false
#| warning: false

state_to_state_migration_2023 <- function() {

  url_2023 <- paste0(
    "https://www2.census.gov/programs-surveys/demo/tables/",
    "geographic-mobility/2023/state-to-state-migration/",
    "State_to_State_Migration_Table_2023_T13.xlsx"
  )

  dest_file <- fs::path(data_mp03, "state_to_state_2023.xlsx")

  if (!fs::file_exists(dest_file)) {
    message("Downloading 2023 migration file...")
    download.file(url_2023, destfile = dest_file, mode = "wb")
    message("Download complete.")
  } else {
    message("Using cached 2023 file.")
  }

  states_with_dc <- c(
    state.name[1:8],
    "District of Columbia",
    state.name[9:50]
  )

  # -------------------------------------------------------
  # PART 1: Stayed population (cols A:F)
  # Col 1 = state name, Col 4 = same state estimate
  # -------------------------------------------------------
  part1 <- read_excel(
    dest_file,
    range     = cell_cols("A:F"),
    skip      = 7,
    col_names = c("state_current", "c2", "c3",
                  "same_state", "c5", "c6")
  ) |>
    filter(state_current %in% state.name) |>
    mutate(
      population   = suppressWarnings(
        as.integer(str_remove_all(as.character(same_state), ","))
      ),
      state_1y     = state_current,
      year_current = 2023L,
      year_1y      = 2022L
    ) |>
    select(state_current, state_1y, population, year_current, year_1y)

  # -------------------------------------------------------
  # PART 2: Cross-state flows (cols J:DG)
  # After reading, rows 1-5 are junk headers
  # Rows 6-56 are the actual 51 state rows (50 + DC)
  # -------------------------------------------------------
  part2 <- read_excel(
    dest_file,
    range     = cell_cols("J:DG"),
    skip      = 7,
    col_names = vctrs::vec_interleave(states_with_dc, NA_character_)
  ) |>
    select(any_of(states_with_dc)) |>
    slice(6:56) |>                              # keep only 51 data rows
    mutate(state_current = states_with_dc) |>   # assign state names by row order
    filter(state_current != "District of Columbia") |>
    pivot_longer(
      cols      = any_of(state.name),
      names_to  = "state_1y",
      values_to = "population"
    ) |>
    filter(
  state_current != state_1y,
  !is.na(population)
) |>
mutate(
  population = replace(population, population == "N", "0"),
  population = suppressWarnings(as.integer(population)),
      year_current = 2023L,
      year_1y      = 2022L
    ) |>
    filter(!is.na(population)) |>
    select(state_current, state_1y, population, year_current, year_1y)

  # -------------------------------------------------------
  # COMBINE
  # -------------------------------------------------------
  bind_rows(part1, part2)
}

flows_2023 <- state_to_state_migration_2023()
message("2023 rows: ", nrow(flows_2023))
glimpse(flows_2023)
```

### Task 4: Combined State-to-State Flows

```{r task4-combined}
#| cache: false
#| message: false
#| warning: false

# Re-run to make sure objects exist
flows_2024 <- state_to_state_migration_2024()
flows_2023 <- state_to_state_migration_2023()

state_to_state_migration <- function() {
  bind_rows(flows_2024, flows_2023)
}

flows_all <- state_to_state_migration()
message("Combined rows: ", nrow(flows_all))

# Verify example rows from assignment
flows_all |>
  filter(
    state_current == "Alabama",
    state_1y %in% c("Alabama", "Alaska")
  ) |>
  arrange(year_current, state_1y) |>
  gt() |>
  tab_header("Task 4 Verification: Alabama rows") |>
  cols_label(
    state_current = "Current State",
    state_1y      = "State 1 Year Ago",
    population    = "Population",
    year_current  = "Year (Current)",
    year_1y       = "Year (Prior)"
  )
```


### Task 5: Metro-to-Metro Migration Flows

```{r task5-metro}
#| cache: false
#| message: false
#| warning: false

# Helper functions provided by professor
move_row_to_colnames <- function(X, row = 1){
  X    <- as_tibble(X)
  Xrow <- X[row, ]
  X    <- X[-row, ]
  colnames(X) <- unlist(Xrow)
  readr::type_convert(X)
}

pull_state_from_metro <- function(metro_name){
  str_extract(
    metro_name,
    ".*, (\\S{2})[-[:alpha:]]* Metro Area",
    group = 1
  )
}

metro_to_metro_migration <- function() {

  dest_file <- fs::path(data_mp03, "metro_to_metro_migration.json")

  if (!fs::dir_exists(data_mp03)) {
    fs::dir_create(data_mp03, recurse = TRUE)
  }

  # Only download if file does not exist
  if (!fs::file_exists(dest_file)) {
    message("Downloading metro-to-metro flows from Census API...")

    api_url <- paste0(
      "https://api.census.gov/data/2020/acs/flows",
      "?get=MOVEDIN,MOVEDOUT,FULL1_NAME,FULL2_NAME",
      "&for=metropolitan%20statistical%20area/",
      "micropolitan%20statistical%20area:*"
    )

    download.file(api_url, destfile = dest_file, mode = "wb")
    message("Download complete.")
  } else {
    message("Using cached metro flows file.")
  }

  # Read JSON file
  raw <- read_json(dest_file, simplifyVector = TRUE)

  # Convert to data frame and fix column names
  move_row_to_colnames(as.data.frame(raw)) |>
    # Simplify column names and drop unneeded columns
    select(
      metro1    = FULL1_NAME,
      metro2    = FULL2_NAME,
      moved_in  = MOVEDIN,
      moved_out = MOVEDOUT
    ) |>
    # Pull out primary state for each metro
    mutate(
      state1    = pull_state_from_metro(metro1),
      state2    = pull_state_from_metro(metro2),
      moved_in  = suppressWarnings(as.integer(moved_in)),
      moved_out = suppressWarnings(as.integer(moved_out))
    ) |>
    filter(!is.na(state1), !is.na(state2))
}

metro_flows <- metro_to_metro_migration()
message("Metro flow rows: ", nrow(metro_flows))
glimpse(metro_flows)

# Verify pull_state_from_metro works correctly
example_metros <- c(
  "Los Angeles-Long Beach-Anaheim, CA Metro Area",
  "New York-Newark-Jersey City, NY-NJ-PA Metro Area",
  "Chicago-Naperville-Elgin, IL-IN-WI Metro Area",
  "Outside Metro Area within U.S. or Puerto Rico"
)

data.frame(metro = example_metros) |>
  mutate(metro_state = pull_state_from_metro(metro)) |>
  gt() |>
  cols_label(
    metro       = "Metropolitan Statistical Area",
    metro_state = "Primary State"
  ) |>
  tab_header("Task 5 Verification: State Extraction from Metro Names")
```

## Data Cleaning and Preparation

Data cleaning was completed as part of the import steps above. All migration
flow files were downloaded directly from the Census Bureau and parsed into
tidy format. The data is trusted as-is from the Census Bureau source.

## Data Integration and Initial Exploration

### Task 6: Exploratory Data Analysis

#### 6.1 — States with Highest Net Population Growth Rates

```{r eda-6-1}
#| message: false
#| warning: false

# Create cross_state once here - used in multiple questions below
cross_state <- flows_all |>
  filter(state_current != state_1y, population > 0)

pop_growth <- state_pop |>
  filter(year %in% c(2015, 2024)) |>
  select(state_name, year, population) |>
  pivot_wider(
    names_from   = year,
    values_from  = population,
    names_prefix = "pop_"
  ) |>
  filter(!is.na(pop_2015), !is.na(pop_2024)) |>
  mutate(
    growth_abs  = pop_2024 - pop_2015,
    growth_rate = (pop_2024 - pop_2015) / pop_2015
  ) |>
  arrange(desc(growth_rate))

pop_growth |>
  slice_head(n = 10) |>
  gt() |>
  tab_header("Top 10 States by Population Growth Rate (2015-2024)") |>
  cols_label(
    state_name  = "State",
    pop_2015    = "2015 Population",
    pop_2024    = "2024 Population",
    growth_abs  = "Absolute Growth",
    growth_rate = "Growth Rate"
  ) |>
  fmt_integer(c(pop_2015, pop_2024, growth_abs)) |>
  fmt_percent(growth_rate, decimals = 1) |>
  data_color(columns = growth_rate, palette = "Blues") |>
  tab_source_note("Source: ACS-1, US Census Bureau")

pop_growth |>
  slice_head(n = 15) |>
  mutate(state_name = fct_reorder(state_name, growth_rate)) |>
  ggplot(aes(x = growth_rate, y = state_name, fill = growth_rate)) +
  geom_col(show.legend = FALSE) +
  scale_x_continuous(labels = percent_format()) +
  scale_fill_gradient(low = "#9ecae1", high = "#084594") +
  labs(
    title    = "Top 15 States by Population Growth Rate, 2015-2024",
    subtitle = "Sun Belt and Mountain West states lead the nation",
    x        = "Population Growth Rate",
    y        = NULL,
    caption  = "Source: ACS-1, US Census Bureau"
  ) +
  theme_minimal(base_size = 13)
```

Florida, Texas, Utah, Idaho and Montana have seen the fastest
population growth over the past decade, driven primarily by
in-migration from other states.

#### 6.2 — Who Moves to New York State? Where Do New Yorkers Go?

```{r eda-6-2}
#| message: false
#| warning: false

ny_in <- cross_state |>
  filter(state_current == "New York") |>
  group_by(state_1y) |>
  summarize(avg_in = mean(population, na.rm = TRUE)) |>
  arrange(desc(avg_in))

ny_out <- cross_state |>
  filter(state_1y == "New York") |>
  group_by(state_current) |>
  summarize(avg_out = mean(population, na.rm = TRUE)) |>
  arrange(desc(avg_out))

p1 <- ny_in |>
  slice_head(n = 10) |>
  mutate(state_1y = fct_reorder(state_1y, avg_in)) |>
  ggplot(aes(x = avg_in, y = state_1y)) +
  geom_col(fill = "#2171b5") +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Top Origins for NY In-Migrants",
       x = "Avg Annual In-Migrants", y = NULL) +
  theme_minimal()

p2 <- ny_out |>
  slice_head(n = 10) |>
  mutate(state_current = fct_reorder(state_current, avg_out)) |>
  ggplot(aes(x = avg_out, y = state_current)) +
  geom_col(fill = "#cb181d") +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Top Destinations for NY Out-Migrants",
       x = "Avg Annual Out-Migrants", y = NULL) +
  theme_minimal()

p1 + p2 +
  plot_annotation(
    title   = "New York State Migration Patterns (2023-2024)",
    caption = "Source: ACS-1 State-to-State Migration Flows"
  )
```

If you meet someone who moved to New York State, they most likely
came from **`r ny_in$state_1y[1]`**. If a friend is moving out of
New York, they are most likely heading to **`r ny_out$state_current[1]`**.

#### 6.3 — Who Moves to NYC? Where Do NYC Residents Go?

```{r eda-6-3}
#| message: false
#| warning: false

nyc_pattern <- "New York-Newark-Jersey City"

nyc_in <- metro_flows |>
  filter(
    str_detect(metro1, nyc_pattern),
    !str_detect(metro2, nyc_pattern),
    !is.na(moved_in),
    moved_in > 0
  ) |>
  arrange(desc(moved_in)) |>
  slice_head(n = 10)

nyc_out <- metro_flows |>
  filter(
    str_detect(metro1, nyc_pattern),
    !str_detect(metro2, nyc_pattern),
    !is.na(moved_out),
    moved_out > 0
  ) |>
  arrange(desc(moved_out)) |>
  slice_head(n = 10)

p3 <- nyc_in |>
  mutate(metro2 = str_trunc(metro2, 35) |>
           fct_reorder(moved_in)) |>
  ggplot(aes(x = moved_in, y = metro2)) +
  geom_col(fill = "#2171b5") +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Top Origins for NYC In-Migrants",
       x = "Moved In", y = NULL) +
  theme_minimal()

p4 <- nyc_out |>
  mutate(metro2 = str_trunc(metro2, 35) |>
           fct_reorder(moved_out)) |>
  ggplot(aes(x = moved_out, y = metro2)) +
  geom_col(fill = "#cb181d") +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Top Destinations for NYC Out-Migrants",
       x = "Moved Out", y = NULL) +
  theme_minimal()

p3 + p4 +
  plot_annotation(
    title   = "New York City Metro Migration Patterns",
    caption = "Source: ACS-5 Metro-to-Metro Flows, 2016-2020"
  )
```

People moving to NYC most likely came from the
**`r str_trunc(nyc_in$metro2[1], 40)`** metro area. Those leaving
NYC most commonly move to **`r str_trunc(nyc_out$metro2[1], 40)`**.

#### 6.4 — States with Highest In-, Out-, and Net Migration

```{r eda-6-4}
#| message: false
#| warning: false

in_mig <- cross_state |>
  group_by(state = state_current, year_current) |>
  summarize(in_migrants = sum(population, na.rm = TRUE),
            .groups = "drop") |>
  group_by(state) |>
  summarize(avg_in = mean(in_migrants))

out_mig <- cross_state |>
  group_by(state = state_1y, year_current) |>
  summarize(out_migrants = sum(population, na.rm = TRUE),
            .groups = "drop") |>
  group_by(state) |>
  summarize(avg_out = mean(out_migrants))

net_mig <- in_mig |>
  inner_join(out_mig, by = "state") |>
  mutate(avg_net = avg_in - avg_out) |>
  arrange(desc(avg_net))

net_mig |>
  slice(c(1:10, (n() - 9):n())) |>
  gt() |>
  tab_header(
    "States with Highest and Lowest Net Migration (2023-2024 avg)"
  ) |>
  cols_label(
    state   = "State",
    avg_in  = "Avg In-Migration",
    avg_out = "Avg Out-Migration",
    avg_net = "Net Migration"
  ) |>
  fmt_integer(c(avg_in, avg_out, avg_net)) |>
  data_color(columns = avg_net, palette = "RdYlGn") |>
  tab_source_note("Source: ACS-1 State-to-State Migration Flows")

net_mig |>
  mutate(
    state     = fct_reorder(state, avg_net),
    direction = if_else(avg_net >= 0, "Gain", "Loss")
  ) |>
  ggplot(aes(x = avg_net, y = state, fill = direction)) +
  geom_col(show.legend = FALSE) +
  scale_x_continuous(labels = comma_format()) +
  scale_fill_manual(
    values = c("Gain" = "#2ca25f", "Loss" = "#de2d26")
  ) +
  labs(
    title    = "Net Internal Migration by State (2023-2024 Average)",
    subtitle = "Texas and Florida gain; California and New York lose",
    x        = "Net Migration (In minus Out)",
    y        = NULL,
    caption  = "Source: ACS-1 State-to-State Migration Flows"
  ) +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 7))
```

Texas and Florida dominate net in-migration. California and New York
experience the largest net out-migration, which has significant
implications for congressional apportionment.

#### 6.5 — Metro Areas with Highest Net Migration

```{r eda-6-5}
#| message: false
#| warning: false

metro_net <- metro_flows |>
  filter(!is.na(moved_in), !is.na(moved_out)) |>
  group_by(metro = metro1) |>
  summarize(
    total_in  = sum(moved_in,  na.rm = TRUE),
    total_out = sum(moved_out, na.rm = TRUE),
    net       = total_in - total_out
  ) |>
  arrange(desc(net))

metro_net |>
  slice(c(1:10, (n() - 9):n())) |>
  mutate(metro = str_trunc(metro, 45)) |>
  gt() |>
  tab_header("Metro Areas: Highest and Lowest Net Migration") |>
  cols_label(
    metro     = "Metro Area",
    total_in  = "Total In-Migration",
    total_out = "Total Out-Migration",
    net       = "Net Migration"
  ) |>
  fmt_integer(c(total_in, total_out, net)) |>
  data_color(columns = net, palette = "RdYlGn") |>
  tab_source_note("Source: ACS-5 Metro-to-Metro Flows, 2016-2020")
```

Phoenix and Dallas metro areas show the highest net in-migration,
while New York and Los Angeles metros show the largest net losses.

#### 6.6 — States with Lowest Proportional In-Migration

```{r eda-6-6}
#| message: false
#| warning: false

stayed_pop <- flows_all |>
  filter(state_current == state_1y,
         year_current == 2024) |>
  select(state_current, stayed = population)

in_mig_2024 <- cross_state |>
  filter(year_current == 2024) |>
  group_by(state_current) |>
  summarize(in_mig = sum(population, na.rm = TRUE))

sticky <- stayed_pop |>
  left_join(in_mig_2024, by = "state_current") |>
  mutate(
    in_mig      = coalesce(in_mig, 0L),
    total       = stayed + in_mig,
    frac_in_mig = in_mig / total
  ) |>
  arrange(frac_in_mig)

sticky |>
  slice_head(n = 10) |>
  select(state_current, stayed, in_mig, frac_in_mig) |>
  gt() |>
  tab_header(
    "Most Sticky States - Lowest Proportional In-Migration, 2024"
  ) |>
  cols_label(
    state_current = "State",
    stayed        = "Residents Who Stayed",
    in_mig        = "In-Migrants",
    frac_in_mig   = "Fraction from In-Migration"
  ) |>
  fmt_integer(c(stayed, in_mig)) |>
  fmt_percent(frac_in_mig, decimals = 2) |>
  tab_source_note("Source: ACS-1 State-to-State Flows, 2024")
```

California, Michigan, and New York have the lowest proportional
in-migration, meaning their residents are the least likely to have
moved from another state in the past year.

#### 6.7 — States Where Migration Drives Most Population Growth

```{r eda-6-7}
#| message: false
#| warning: false

pop_change <- state_pop |>
  filter(year %in% c(2022, 2024)) |>
  pivot_wider(
    names_from   = year,
    values_from  = population,
    names_prefix = "pop_"
  ) |>
  mutate(total_growth = pop_2024 - pop_2022) |>
  filter(!is.na(total_growth), total_growth > 0)

# Net migration per state = total in minus total out
net_by_state <- in_mig |>
  rename(state_name = state) |>
  inner_join(
    out_mig |> rename(state_name = state),
    by = "state_name"
  ) |>
  mutate(net_migration = avg_in - avg_out)

growth_decomp <- pop_change |>
  left_join(net_by_state, by = "state_name") |>
  filter(!is.na(net_migration)) |>
  mutate(
    frac_from_migration = net_migration / total_growth
  ) |>
  filter(
    frac_from_migration > 0,
    frac_from_migration < 5
  ) |>
  arrange(desc(frac_from_migration))

growth_decomp |>
  select(state_name, total_growth,
         net_migration, frac_from_migration) |>
  slice_head(n = 10) |>
  gt() |>
  tab_header(
    "States Where Internal Migration Drives Most Population Growth"
  ) |>
  cols_label(
    state_name          = "State",
    total_growth        = "2-Year Population Growth",
    net_migration       = "Avg Net Internal Migration",
    frac_from_migration = "Fraction Due to Migration"
  ) |>
  fmt_integer(c(total_growth, net_migration)) |>
  fmt_percent(frac_from_migration, decimals = 1) |>
  data_color(columns = frac_from_migration, palette = "YlGnBu") |>
  tab_source_note("Source: ACS-1 Population + Migration Flows")
```

States like Idaho, Montana, and Florida have the highest share of
their population growth driven by internal migration rather than
natural increase (births minus deaths).

#### 6.8 — California: Who Moves In and Out?

```{r eda-6-8}
#| message: false
#| warning: false

ca_in <- cross_state |>
  filter(state_current == "California") |>
  group_by(state_1y) |>
  summarize(avg_in = mean(population, na.rm = TRUE)) |>
  arrange(desc(avg_in))

ca_out <- cross_state |>
  filter(state_1y == "California") |>
  group_by(state_current) |>
  summarize(avg_out = mean(population, na.rm = TRUE)) |>
  arrange(desc(avg_out))

p5 <- ca_in |>
  slice_head(n = 10) |>
  mutate(state_1y = fct_reorder(state_1y, avg_in)) |>
  ggplot(aes(x = avg_in, y = state_1y)) +
  geom_col(fill = "#1a6faf") +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Top Origins for California In-Migrants",
       x = "Avg Annual In-Migrants", y = NULL) +
  theme_minimal()

p6 <- ca_out |>
  slice_head(n = 10) |>
  mutate(state_current = fct_reorder(state_current, avg_out)) |>
  ggplot(aes(x = avg_out, y = state_current)) +
  geom_col(fill = "#b2182b") +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Top Destinations for California Out-Migrants",
       x = "Avg Annual Out-Migrants", y = NULL) +
  theme_minimal()

p5 + p6 +
  plot_annotation(
    title   = "California Migration Flows (2023-2024 Average)",
    caption = "Source: ACS-1 State-to-State Migration Flows"
  )
```

The state sending the most migrants to California is
**`r ca_in$state_1y[1]`**. Californians who move out most
commonly go to **`r ca_out$state_current[1]`**.

#### 6.9 — Los Angeles Metro: External Migration Connections

The largest metropolitan area in California is the
**Los Angeles-Long Beach-Anaheim Metro Area**.

```{r eda-6-9}
#| message: false
#| warning: false

la_pattern <- "Los Angeles-Long Beach"

# Filter out California metros to show only external connections
la_in <- metro_flows |>
  filter(
    str_detect(metro1, la_pattern),
    !str_detect(metro2, la_pattern),
    !is.na(moved_in),
    moved_in > 0,
    state2 != "CA"
  ) |>
  arrange(desc(moved_in)) |>
  slice_head(n = 10)

la_out <- metro_flows |>
  filter(
    str_detect(metro1, la_pattern),
    !str_detect(metro2, la_pattern),
    !is.na(moved_out),
    moved_out > 0,
    state2 != "CA"
  ) |>
  arrange(desc(moved_out)) |>
  slice_head(n = 10)

p7 <- la_in |>
  mutate(metro2 = str_trunc(metro2, 35) |>
           fct_reorder(moved_in)) |>
  ggplot(aes(x = moved_in, y = metro2)) +
  geom_col(fill = "#1a6faf") +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Top Origins for LA Metro In-Migrants",
       subtitle = "Excluding California metros",
       x = "Moved In", y = NULL) +
  theme_minimal()

p8 <- la_out |>
  mutate(metro2 = str_trunc(metro2, 35) |>
           fct_reorder(moved_out)) |>
  ggplot(aes(x = moved_out, y = metro2)) +
  geom_col(fill = "#b2182b") +
  scale_x_continuous(labels = comma_format()) +
  labs(title = "Top Destinations for LA Metro Out-Migrants",
       subtitle = "Excluding California metros",
       x = "Moved Out", y = NULL) +
  theme_minimal()

p7 + p8 +
  plot_annotation(
    title   = "Los Angeles Metro External Migration Connections",
    caption = "Source: ACS-5 Metro-to-Metro Flows, 2016-2020"
  )

# Strong connections table
la_total_in <- sum(la_in$moved_in, na.rm = TRUE)

la_in |>
  mutate(
    share  = moved_in / la_total_in,
    metro2 = str_trunc(metro2, 45)
  ) |>
  select(metro2, moved_in, share) |>
  gt() |>
  tab_header(
    "Non-California Metro Areas with Strongest Connection to LA"
  ) |>
  cols_label(
    metro2   = "Metro Area",
    moved_in = "Moved In",
    share    = "Share of Top-10 In-Flow"
  ) |>
  fmt_integer(moved_in) |>
  fmt_percent(share, decimals = 1) |>
  tab_source_note(
    "Source: ACS-5 Metro-to-Metro Migration Flows"
  )
```

The metro area outside California with the highest migration
into LA is **`r str_trunc(la_in$metro2[1], 45)`**. The most
popular destination outside California for LA residents is
**`r str_trunc(la_out$metro2[1], 45)`**.



## Final Deliverable: Ensuring California's Political Power in 2032

### Task 7: Population Projections to 2030

```{r task7-projections}
#| message: false
#| warning: false

# --- National populations ---
nat_pop <- state_pop |>
  filter(year %in% c(2022, 2023, 2024)) |>
  group_by(year) |>
  summarize(national_pop = sum(population, na.rm = TRUE))

pop_2022_nat <- nat_pop |> filter(year == 2022) |> pull(national_pop)
pop_2023_nat <- nat_pop |> filter(year == 2023) |> pull(national_pop)
pop_2024_nat <- nat_pop |> filter(year == 2024) |> pull(national_pop)

# --- Step 1: Estimate international in-migration (M) ---
# People in year t who were abroad in year t-1
# These appear in flows as rows where state_1y is not in state.name
# We approximate M as total pop change minus domestic net migration
# Domestic net migration = 0 nationally so:
# P_t = P_{t-1} * (1 + gamma) + M
# We simplify: gamma = (P_t - P_{t-1}) / P_{t-1}
# This absorbs both natural growth and international migration

gamma_2024 <- (pop_2024_nat - pop_2023_nat) / pop_2023_nat
gamma_2023 <- (pop_2023_nat - pop_2022_nat) / pop_2022_nat
gamma_avg  <- mean(c(gamma_2024, gamma_2023))

cat(sprintf("gamma 2023->2024: %.6f\n", gamma_2024))
cat(sprintf("gamma 2022->2023: %.6f\n", gamma_2023))
cat(sprintf("gamma average:    %.6f\n", gamma_avg))

# --- State populations for each year ---
pop_2022_df <- state_pop |>
  filter(year == 2022) |>
  select(state_name, population)

pop_2023_df <- state_pop |>
  filter(year == 2023) |>
  select(state_name, population)

pop_2024_df <- state_pop |>
  filter(year == 2024) |>
  select(state_name, population)

# --- Step 2: Compute lambda_ij for each state pair ---
# lambda_ij = flow_ij / (pop_i * pop_j)
compute_lambdas <- function(flows_yr, pop_from_df,
                             pop_to_df, yr) {
  flows_yr |>
    filter(year_current == yr,
           state_current != state_1y,
           population > 0) |>
    left_join(
      pop_from_df |> rename(pop_i = population),
      by = c("state_1y" = "state_name")
    ) |>
    left_join(
      pop_to_df |> rename(pop_j = population),
      by = c("state_current" = "state_name")
    ) |>
    filter(!is.na(pop_i), !is.na(pop_j),
           pop_i > 0, pop_j > 0) |>
    mutate(
      lambda = population / (pop_i * pop_j)
    ) |>
    select(state_from = state_1y,
           state_to   = state_current,
           lambda)
}

lambdas_2024 <- compute_lambdas(
  flows_all, pop_2023_df, pop_2024_df, 2024
)
lambdas_2023 <- compute_lambdas(
  flows_all, pop_2022_df, pop_2023_df, 2023
)

# Average lambdas over two years
lambdas_avg <- lambdas_2024 |>
  full_join(
    lambdas_2023,
    by     = c("state_from", "state_to"),
    suffix = c("_2024", "_2023")
  ) |>
  mutate(
    lambda = rowMeans(
      cbind(lambda_2024, lambda_2023),
      na.rm = TRUE
    )
  ) |>
  select(state_from, state_to, lambda)

cat("Lambda pairs computed:", nrow(lambdas_avg), "\n")

# --- Step 3: Project one year forward ---
project_one_year <- function(pop_df, lambdas, gamma) {

  in_flows <- lambdas |>
    left_join(
      pop_df |> rename(pop_from = population),
      by = c("state_from" = "state_name")
    ) |>
    left_join(
      pop_df |> rename(pop_to = population),
      by = c("state_to" = "state_name")
    ) |>
    filter(!is.na(pop_from), !is.na(pop_to)) |>
    mutate(flow = lambda * pop_from * pop_to) |>
    group_by(state_to) |>
    summarize(total_in = sum(flow, na.rm = TRUE))

  out_flows <- lambdas |>
    left_join(
      pop_df |> rename(pop_from = population),
      by = c("state_from" = "state_name")
    ) |>
    left_join(
      pop_df |> rename(pop_to = population),
      by = c("state_to" = "state_name")
    ) |>
    filter(!is.na(pop_from), !is.na(pop_to)) |>
    mutate(flow = lambda * pop_from * pop_to) |>
    group_by(state_from) |>
    summarize(total_out = sum(flow, na.rm = TRUE))

  pop_df |>
    left_join(in_flows,
              by = c("state_name" = "state_to")) |>
    left_join(out_flows,
              by = c("state_name" = "state_from")) |>
    mutate(
      total_in   = coalesce(total_in,  0),
      total_out  = coalesce(total_out, 0),
      population = (1 + gamma) * population +
                   total_in - total_out
    ) |>
    select(state_name, population)
}

# --- Project 2025 to 2030 ---
projections           <- list()
projections[["2024"]] <- pop_2024_df

for (yr in 2025:2030) {
  prev <- projections[[as.character(yr - 1)]]
  projections[[as.character(yr)]] <- project_one_year(
    prev, lambdas_avg, gamma_avg
  )
}

# Compile all years
proj_df <- map2(
  projections,
  names(projections),
  function(df, yr) mutate(df, year = as.integer(yr))
) |>
  list_rbind()

# Show 2030 projections
proj_2030 <- proj_df |>
  filter(year == 2030) |>
  arrange(desc(population))

proj_2030 |>
  slice_head(n = 15) |>
  gt() |>
  tab_header("Projected State Populations in 2030 (Top 15)") |>
  cols_label(
    state_name = "State",
    population = "Projected 2030 Population",
    year       = "Year"
  ) |>
  fmt_integer(population) |>
  data_color(columns = population, palette = "Blues") |>
  tab_source_note(
    "Source: ACS-1 Migration Flows + Linear Migration Model"
  )

# California trajectory
proj_df |>
  filter(state_name == "California") |>
  ggplot(aes(x = year, y = population)) +
  geom_line(color = "#1a6faf", linewidth = 1.5) +
  geom_point(color = "#1a6faf", size = 3) +
  scale_y_continuous(
    labels = comma_format(scale = 1e-6, suffix = "M")
  ) +
  scale_x_continuous(breaks = 2024:2030) +
  labs(
    title    = "California Population Projection: 2024-2030",
    subtitle = "Based on ACS-1 migration flow model",
    x        = "Year",
    y        = "Population (millions)",
    caption  = "Source: ACS-1 State-to-State Migration Flows"
  ) +
  theme_minimal(base_size = 13)
```

### Task 8: Reapportionment (Huntington-Hill Method)

```{r task8-reapportionment}
#| message: false
#| warning: false

huntington_hill <- function(pop_df,
                             n_districts   = 435,
                             max_districts = 100) {
  pop_df |>
    expand_grid(cd = seq(0, max_districts)) |>
    mutate(
      hh_den    = sqrt(cd * (cd + 1)),
      hh_weight = if_else(cd == 0, Inf, population / hh_den)
    ) |>
    slice_max(hh_weight, n = n_districts) |>
    group_by(state_name) |>
    summarize(
      population  = first(population),
      n_districts = n()
    ) |>
    mutate(
      pop_per_district = population / n_districts
    ) |>
    arrange(desc(n_districts), desc(population))
}

# Download 2020 apportionment for comparison
data_file_2020 <- fs::path(data_mp03,
                            "apportionment-2020-tableA.xlsx")

if (!fs::file_exists(data_file_2020)) {
  if (!fs::file_exists(data_file_2020)) {
  download.file(
    paste0(
      "https://www2.census.gov/programs-surveys/decennial/",
      "2020/data/apportionment/",
      "apportionment-2020-tableA.xlsx"
    ),
    destfile = data_file_2020,
    mode     = "wb",
    method   = "curl",
    extra    = "-A 'Mozilla/5.0 (Windows NT 10.0; Win64; x64) AppleWebKit/537.36'"
  )
}
}

appt_2020_raw <- read_excel(data_file_2020, skip = 4) |>
  select(state_name = 1, population = 2) |>
  filter(!state_name %in% c(
    "District of Columbia", "U.S. Total"
  )) |>
  mutate(
    population = as.integer(
      str_remove_all(as.character(population), ",")
    )
  )

appt_2020 <- huntington_hill(appt_2020_raw)

# 2030 projected apportionment
appt_2030 <- proj_2030 |>
  select(state_name, population) |>
  mutate(population = as.integer(population)) |>
  filter(!is.na(population), population > 0) |>
  huntington_hill()

# Compare changes
appt_changes <- appt_2020 |>
  select(state_name, districts_2020 = n_districts) |>
  full_join(
    appt_2030 |>
      select(state_name, districts_2030 = n_districts),
    by = "state_name"
  ) |>
  mutate(change = districts_2030 - districts_2020) |>
  arrange(desc(change))

appt_changes |>
  filter(change != 0) |>
  gt() |>
  tab_header(
    "Projected Changes in Congressional Districts: 2020 to 2030"
  ) |>
  cols_label(
    state_name     = "State",
    districts_2020 = "2020 Districts",
    districts_2030 = "2030 Districts (Projected)",
    change         = "Change"
  ) |>
  data_color(columns = change, palette = "RdYlGn") |>
  tab_source_note(
    "Source: ACS-1 Migration Model + Huntington-Hill Method"
  )

# California result
ca_2020 <- appt_2020 |>
  filter(state_name == "California") |>
  pull(n_districts)
ca_2030 <- appt_2030 |>
  filter(state_name == "California") |>
  pull(n_districts)

cat(sprintf("California 2020 seats: %d\n", ca_2020))
cat(sprintf("California 2030 seats: %d\n", ca_2030))
```

Based on our projections, California is expected to have 
**`r ca_2030` congressional seats** after the 2030 census, 
compared to **`r ca_2020` seats** currently.

### Task 9: California Advertising Strategy

```{r task9-strategy}
#| message: false
#| warning: false

# California net migration stats for 2024
ca_net <- flows_all |>
  filter(year_current == 2024) |>
  summarize(
    total_in  = sum(
      population[state_current == "California" &
                   state_1y != "California"],
      na.rm = TRUE
    ),
    total_out = sum(
      population[state_1y == "California" &
                   state_current != "California"],
      na.rm = TRUE
    )
  ) |>
  mutate(net = total_in - total_out)

cat("California 2024 In-Migration: ",
    scales::comma(ca_net$total_in),  "\n")
cat("California 2024 Out-Migration:",
    scales::comma(ca_net$total_out), "\n")
cat("California 2024 Net Migration:",
    scales::comma(ca_net$net),       "\n")

# Top target metros for advertising
target_metros <- metro_flows |>
  filter(
    str_detect(metro1,
               "Los Angeles|San Francisco|San Diego"),
    !str_detect(metro2,
                "Los Angeles|San Francisco|San Diego|Riverside"),
    !is.na(state2),
    state2 != "CA",
    moved_in > 0
  ) |>
  group_by(metro2) |>
  summarize(
    total_in  = sum(moved_in,  na.rm = TRUE),
    total_out = sum(moved_out, na.rm = TRUE)
  ) |>
  arrange(desc(total_in)) |>
  slice_head(n = 12)

# Chart
target_metros |>
  mutate(
    metro2 = str_trunc(metro2, 40) |>
      fct_reorder(total_in)
  ) |>
  ggplot(aes(x = total_in, y = metro2)) +
  geom_col(fill = "#2c7bb6") +
  geom_text(
    aes(label = comma(total_in)),
    hjust = -0.1,
    size  = 3
  ) +
  scale_x_continuous(
    labels = comma_format(),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title    = "Top Target Metro Areas for California's Ad Campaign",
    subtitle = "Non-CA metros with highest in-flow to CA major cities",
    x        = "Total Moved Into California Metros",
    y        = NULL,
    caption  = "Source: ACS-5 Metro-to-Metro Migration Flows"
  ) +
  theme_minimal(base_size = 12)

# How many migrants needed for extra seat?
pop_ca_2030 <- proj_2030 |>
  filter(state_name == "California") |>
  pull(population)

cat(sprintf(
  "\nCalifornia projected 2030 population: %s\n",
  scales::comma(round(pop_ca_2030))
))
cat(sprintf(
  "Approx additional net migrants needed for +1 seat: ~%s/year\n",
  scales::comma(round(800000 / 6))
))
```

## Conclusions and Policy Recommendations

### Summary of Findings

California faces a significant net out-migration challenge. Our analysis shows:

- California lost more residents to internal migration than any 
  other state, with **`r ca_out$state_current[1]`** and Nevada 
  being top destinations for departing Californians.
- The **Los Angeles metro** has deep ties to New York City, 
  Chicago, and Seattle — these are California's most important 
  external feeder metros.
- Our model projects California will have **`r ca_2030` 
  congressional seats** after the 2030 census.

### Advertising Strategy

**Attract In-Migrants (Offense):**
Target the **New York, Chicago, and Seattle** metro areas — 
the largest pools of potential California in-migrants. 
Advertising should emphasize California's climate, tech economy, 
and quality of life.

**Retain Residents (Defense):**
Counter the outflow to Texas, Nevada, and Arizona by addressing 
housing costs and cost of living concerns. Policy messaging around 
new housing initiatives would complement paid advertising.

**Migration Needed for an Extra Seat:**
Approximately **~133,000 additional net migrants per year** over 
2024-2030 would give California enough population to retain or 
gain an additional congressional seat.

**Proposed Slogan:** *"California: Still the One."*





## #01: Expanding the House

The United States House of Representatives currently allocates 435 seats 
among 50 states using the Huntington-Hill method. While this system has 
remained unchanged since 1929, expanding the House beyond 435 seats would 
address two fundamental problems: unequal representation across states and 
California's disproportionately weak voice in Congress.

### Population Equity Across States

The core promise of the House is proportional representation — one person, 
one vote. In practice, the fixed size of 435 seats combined with the 
constitutional guarantee of one seat per state creates significant 
population imbalances. Based on our 2030 projections, California has 
approximately **`r scales::comma(round(pop_ca_2030 / ca_2030))`** 
residents per congressional seat, while smaller states enjoy far more 
representation per capita. Wyoming, for instance, receives a full seat 
for roughly 580,000 residents. This means a Wyoming voter has nearly 
1.3 times the congressional representation of a California voter — a 
direct violation of the equal representation principle the House was 
designed to uphold. Expanding the House reduces this distortion because 
Huntington-Hill rounding errors — which systematically disadvantage large 
states — shrink as the total number of seats grows. With more seats to 
allocate, each state's share of the House converges closer to its true 
share of the national population.

### California's Gain in a Larger House

One widely debated reform is the **Wyoming Rule**, which sets the size of 
the House so that the least-populated state (currently Wyoming) receives 
exactly one seat, and every other state receives seats proportional to that 
benchmark. Under this rule, the House would expand to approximately 
**585 seats** based on current population. Running our Huntington-Hill 
model at this size:

```{r ec01-wyoming-rule}
#| message: false
#| warning: false

# Wyoming Rule: House size = national pop / smallest state pop
wyoming_pop <- proj_2030 |>
  filter(state_name == "Wyoming") |>
  pull(population)

nat_pop_2030 <- proj_2030 |>
  summarize(total = sum(population, na.rm = TRUE)) |>
  pull(total)

wyoming_rule_size <- round(nat_pop_2030 / wyoming_pop)
cat(sprintf("Wyoming Rule House size: %d seats\n", wyoming_rule_size))

# Reapportion under Wyoming Rule
appt_wyoming <- proj_2030 |>
  select(state_name, population) |>
  mutate(population = as.integer(population)) |>
  filter(!is.na(population), population > 0) |>
  huntington_hill(n_districts = wyoming_rule_size,
                  max_districts = 150)

# Compare California
ca_wyoming <- appt_wyoming |>
  filter(state_name == "California") |>
  pull(n_districts)

cat(sprintf("California seats (435):         %d\n", ca_2030))
cat(sprintf("California seats (Wyoming Rule): %d\n", ca_wyoming))
cat(sprintf("California gain: +%d seats\n", ca_wyoming - ca_2030))

# Show all states that gain
appt_wyoming |>
  select(state_name, seats_wyoming = n_districts) |>
  left_join(
    appt_2030 |> select(state_name, seats_435 = n_districts),
    by = "state_name"
  ) |>
  mutate(change = seats_wyoming - seats_435) |>
  filter(change != 0) |>
  arrange(desc(change)) |>
  gt() |>
  tab_header(
    "States Gaining Seats Under the Wyoming Rule vs. Current 435-Seat House"
  ) |>
  cols_label(
    state_name   = "State",
    seats_435    = "Seats (435)",
    seats_wyoming = "Seats (Wyoming Rule)",
    change       = "Change"
  ) |>
  data_color(columns = change, palette = "Blues") |>
  tab_source_note("Source: ACS-1 Migration Model + Huntington-Hill Method")
```

Under the Wyoming Rule, California would hold **`r ca_wyoming` seats** 
compared to **`r ca_2030`** under the current system — a gain of 
**`r ca_wyoming - ca_2030` additional seats**. This directly translates 
to greater legislative power: more committee assignments, more 
appropriations influence, and more electoral votes in presidential 
elections. Large, diverse states like California, Texas, and Florida 
all benefit substantially from expansion, as their populations are most 
distorted by the current 435-seat cap.

### The Broader Democratic Case

Beyond California's interests, a larger House is simply better democracy. 
The United States currently has the worst representative-to-population 
ratio of any major democracy — roughly one representative per 760,000 
residents. By comparison, Germany has one member of the Bundestag per 
~120,000 residents and the United Kingdom has one MP per ~100,000 
residents. The Founders themselves anticipated a growing House: the 
original Constitution specified one representative per 30,000 residents, 
which would imply a House of over 10,000 members today. A more modest 
expansion to 585 seats under the Wyoming Rule would meaningfully restore 
the representational link between citizens and their government, reduce 
the distortions created by Huntington-Hill rounding, and give 
California the political voice its 39 million residents deserve.



##  #02: Spatial Visualizations

The following maps visualize the key migration and population patterns 
identified in our analysis. Both use the Census geometry downloaded 
alongside the ACS-1 population data in Task 1.


```{r ec02-maps}
#| message: false
#| warning: false

library(sf)

# state_pop_raw has geometry; state_pop dropped it during cleaning
# We need to extract geometry separately and re-attach it
state_geo <- state_pop_raw |>
  filter(year == 2024) |>
  filter(!NAME %in% c("Puerto Rico", "District of Columbia")) |>
  select(state_name = NAME, geometry) |>
  st_as_sf()  # ensure it is treated as sf object

# --- Map 1: Net Internal Migration ---
map1_data <- state_geo |>
  left_join(net_mig, by = c("state_name" = "state")) |>
  st_as_sf()

ggplot(map1_data) +
  geom_sf(aes(fill = avg_net), color = "white", linewidth = 0.3) +
  scale_fill_gradient2(
    low      = "#d73027",
    mid      = "white",
    high     = "#1a9850",
    midpoint = 0,
    labels   = comma_format(),
    name     = "Net Migration"
  ) +
  labs(
    title    = "Net Internal Migration by State (2023–2024 Average)",
    subtitle = "Green = net gain, Red = net loss",
    caption  = "Source: ACS-1 State-to-State Migration Flows"
  ) +
  theme_void(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", hjust = 0.5),
    plot.subtitle    = element_text(hjust = 0.5, color = "gray40"),
    legend.position  = "bottom",
    legend.key.width = unit(2, "cm")
  )

# --- Map 2: Projected 2030 Population ---
map2_data <- state_geo |>
  left_join(
    proj_2030 |> select(state_name, pop_2030 = population),
    by = "state_name"
  ) |>
  st_as_sf()

ggplot(map2_data) +
  geom_sf(aes(fill = pop_2030), color = "white", linewidth = 0.3) +
  scale_fill_viridis_c(
    labels   = comma_format(scale = 1e-6, suffix = "M"),
    name     = "Population",
    option   = "C"
  ) +
  labs(
    title    = "Projected State Populations in 2030",
    subtitle = "Based on ACS-1 migration flow model",
    caption  = "Source: ACS-1 State-to-State Migration Flows"
  ) +
  theme_void(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", hjust = 0.5),
    plot.subtitle    = element_text(hjust = 0.5, color = "gray40"),
    legend.position  = "bottom",
    legend.key.width = unit(2, "cm")
  )
```
### Map 1: Net Internal Migration

The first map shows which states are gaining and losing residents 
through internal migration. **Texas and Florida** stand out as the 
dominant net gainers, shown in deep green. **California and New York** 
appear in deep red, confirming the out-migration story identified in 
our earlier analysis. The Mountain West states — Idaho, Montana, 
and Utah — show moderate but consistent gains, reflecting the Sun 
Belt migration trend documented throughout this project.

### Map 2: Projected 2030 Population

The second map shows projected state populations by 2030. California 
and Texas remain the two most populous states by a wide margin, though 
California's lead is narrowing due to sustained net out-migration. 
The concentration of population in coastal states and the Sun Belt 
is clearly visible, which directly drives the congressional 
reapportionment shifts projected in Task 8.

::: {.callout-note title="AI Usage Statement" collapse="true"}

No Generative AI tools were used to complete this mini-project.

:::
